---
title: "Week 12 - Part II - R as a GIS"
author: "Jose Luis Rodriguez Gil"
date: "27/11/2020"
output: 
  html_document:
    number_sections: true
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}

library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(viridis)
library(gt)

library(sf)     # sf will be our main tool to work with spatial data
library(rgdal)  # Another comon package, while on its way out. We will need it for some applications
library(raster) # sf can only handle vector data, we need a special package for raster (drid) data

library(ggspatial)  # for adding scale and north arrow to plots

library(lakemorpho)  # USGS package for lake analysis

```

```{r my theme, include=FALSE}

source(here("functions", "theme_pepe.R"))

```

# Some resources

1. [Geocomputation with R](https://geocompr.robinlovelace.net/)
2. Nice [resource](https://thinking-spatial.org/courses/angewandte_geodatenverarbeitung/kurs06/) about the sf package


# Loading some data

We are going to use the dataset of water quality variables measuerd in Lake winnipeg in 2002-2004 that we used in previous classes:

```{r}

lw_data_original <- read_csv(here("data", "lake_winnipeg_chemistry_2002_2004.csv"))

```


But now we need as well a `.shp` layer with lake Winnipeg. This was extracted from the government of Canada [Water File - Lakes and Rivers (polygons) - 2011 Census](https://open.canada.ca/data/en/dataset/448ec403-6635-456b-8ced-d3ac24143add)

```{r}

lake_winnipeg <- read_sf(here("data", "lw", "lw.shp"))

plot(lake_winnipeg, axes = TRUE)

```


# Manipulating the data

We are going to clean up the water quality data and keep just the total phosphorus data for the surface depth in 2004

```{r}

lw_data <- lw_data_original %>% 
  clean_names() %>% 
  mutate(year = year(activity_start_date)) %>% 
  filter(characteristic_name == "Total Phosphorus, mixed forms") %>% 
  filter(activity_depth_height_measure == 0) %>% 
  filter(year == 2004) %>% 
  dplyr::select(monitoring_location_latitude,
         monitoring_location_longitude,
         result_value)

print(lw_data)

```


# converting a dataframe into a geospatial object with `sf`

We can turn this tibble into a proper geospatial object with `sf`, we just need to specify the **Coordenates Reference System (CRS)**. The data set indicates that it is WG84 (typical for GPS). To simplyfy things we can use the **EPSG code 4326**.

```{r}

lw_data_sf <- st_as_sf(lw_data, coords = c("monitoring_location_longitude", "monitoring_location_latitude"), crs = 4326) # data listed as WGS84


```

WE can check the geospatial information for each of our layers

```{r}

st_crs(lake_winnipeg)

st_geometry(lake_winnipeg)

st_crs(lw_data_sf)

st_geometry(lw_data_sf)

```

The lake layer is in a different CRS, so lets transform it into EPSG 4326 so they both match

```{r}

lw <- st_transform(lake_winnipeg, 4326)

st_crs(lw)

```


# Ploting geospatial data with `ggplot`

```{r}

ggplot() +
  geom_sf(data = lw, 
          fill = "#c6def1", 
          alpha = 0.7) +
  geom_sf(data = lw_data_sf,
          aes(colour = result_value),
          alpha = 0.7,
          show.legend = "point",
          size = 1.1) +
  
  scale_colour_viridis(option = "magma") +
  coord_sf(datum = st_crs(4326)) +
  
  annotation_scale(location = "bl", 
                   width_hint = 0.5,
                   height = unit(0.1, "cm")) +
  annotation_north_arrow(location = "tr", which_north = "true",
                         height = unit(0.6, "cm"),
                         width = unit(0.3, "cm"),
                         # pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
                         style = north_arrow_orienteering(text_size = 5)) +
  theme_pepe() +
  labs(title = "Lake Winnipeg Total Phosphorus",
       subtitle = "Surface total phosphorus - lake Winnipeg - 2004")

```

# Some basic operations

`sf()` allows for many of the common operations that a GIS application would provide (for vector data). For example, we could add a 0.1 degree buffer around the lake, which we could use to, for example assess if speciffic land uses around the shoreline are resposible for the Total Phosphorus concentrations, or to create flood risk maps.

```{r}

lw_buffer <- st_buffer(lake_winnipeg, units::set_units(0.1, degree))

```

We can plot this buffer

```{r}


ggplot() +
  geom_sf(data = lw_buffer, 
          fill = "#f9844a", 
          colour = NA, 
          alpha = 0.4) +
  geom_sf(data = lw, 
          fill = "#c6def1", 
          alpha = 0.7) +
  coord_sf(datum = st_crs(4326)) +
  
  annotation_scale(location = "bl", 
                   width_hint = 0.5,
                   height = unit(0.1, "cm")) +
  annotation_north_arrow(location = "tr", which_north = "true",
                         height = unit(0.6, "cm"),
                         width = unit(0.3, "cm"),
                         # pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
                         style = north_arrow_orienteering(text_size = 5)) +
  
  theme_pepe()
```

# Lakemorpho, a useful tool for lake geospatial analysis

```{r}

lake_260 <- readOGR(dsn = here("data", "lake_260", "L260_outline_milne"), layer = "L260_outline_milne")
lake_260_dem <- raster(here("data", "lake_260","DEM clip.tif"))

```

```{r}

plot(lake_260, axes = TRUE)

```



```{r}
plot(lake_260_dem)

```

```{r message=FALSE, warning=FALSE}

lake_260_morph <- lakeSurroundTopo(lake_260, lake_260_dem, inCatch = NULL, reso = res(lake_260_dem)[1])

plot(lake_260_morph)

```

```{r message=FALSE, warning=FALSE}

fetches_260 <- tibble(bearing = seq(0,360,10)) %>% 
  mutate(fetch = map_dbl(.x = bearing, ~lakeFetch(lake_260_morph,.x))) %>% 
  arrange(desc(fetch))

print(fetches_260)

```

```{r}

Lake_260_metrics <- calcLakeMetrics(lake_260_morph, 40, 250) %>% 
  unlist() %>% 
  as_tibble_row()

print(Lake_260_metrics)

```

```{r}
Lake_260_metrics %>% 
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "measure") %>% 
  mutate(measure = round(measure, digits = 2)) %>% 
  gt()
```







